---
title: "1. Basic Usage"
date: "`r Sys.Date()`"
author: "Nikhil Kaza"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
    code-fold: false
vignette: >
  %\VignetteIndexEntry{1. Basic Usage}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center")
```


This vignette is a practical, code-first tour of the  shape indices in `gridmorph`, a package that provides them for rasters. The indices are conceptually same as the ones in  `shapeindices` (which works on `sf` (multi)polygons).  For reference, here is a recap of the indices

- *convexity*: exterior proportion of lines that connect interior points 
- *moment of inertia*: distribution of shape relative to its center of gravity
- *isotropy*: preference of shape to an axis
- *directional balance*: preference of shape to a direction
- *span*: average interior point distance within the shape
- *radial concentration*: concentration relative to center in the shape
- *depth*: average distance of interior of the shape to the edge 
- *hull ratio*: proportion of the shape to its convex hull
- *polsby popper*: relative area to perimeter of the shape
- *width length ratio*: ratio of the sides of the rotated bounding box of the shape
- *reock*: proportion of the shape area to the minimum bounding circle
- *detour*: ratio of the perimeters of the shape's equal area circle and its convex hull
- *exchange*: proportion of the shape area covered by the equal area circle centered on the shape.

In addition to these, rasters automatically admit a couple more with little difficulty and are thus included. They are ratios of shape constrained shortest path distance to euclidean distance. The domain (choice of points) determine the index

- *geodesic span*: interior points
- *geodesic chord*: perimeter points

For the mathematical grounding and where each index breaks down, see [vignettes in `shapeindices`](https://nkaza.github.io/shapeindices)

In addition, the package also provides six conventional morphological
`erode()``dilate()`, `opening()`, `closing()`, `tophat()`,
`bottomhat()`. However, unlike image processing literature, these operators provided here are sensitive to the raster type (e.g. binary, continuous and categorical)


# Six canonical shapes

Six small synthetic rasters, built directly from cell coordinates, cover
a useful range of shape behaviour: a **disk** (the reference shape every
index treats as optimal), a **pinwheel** (four spiky, rotationally
symmetric blades), a **dumbbell** (two round masses joined by a thin
bar), a **ring** (a disk with its centre hollowed out), an **L-shape**
(a simple right-angle notch), and a **patchy** raster (several disjoint
patches). The first five are single, connected pieces; `patchy` is not.


```{r shapes}
#| code-fold: true

library(gridmorph)
library(terra)

make_disk <- function(n = 81, radius = 30) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  ifel(sqrt(cx^2 + cy^2) <= radius, 1, 0)
}

make_pinwheel <- function(n = 81, blades = 4, r_max = 35) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  theta <- (atan2(cy, cx) + 2 * pi) %% (2 * pi)
  sector_width <- 2 * pi / blades
  theta_local <- theta %% sector_width
  # each blade sweeps from a point (theta_local = 0) out to r_max
  # (theta_local = sector_width) - `blades` copies, evenly rotated
  ifel(sqrt(cx^2 + cy^2) <= r_max * (theta_local / sector_width), 1, 0)
}

make_dumbbell <- function(n = 81, r_ball = 16, bar_half_width = 4, gap = 50) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  ball1 <- sqrt((cx + gap / 2)^2 + cy^2) <= r_ball
  ball2 <- sqrt((cx - gap / 2)^2 + cy^2) <= r_ball
  bar   <- (cx >= -gap / 2) & (cx <= gap / 2) & (abs(cy) <= bar_half_width)
  ifel(ball1 | ball2 | bar, 1, 0)
}

make_ring <- function(n = 81, r_outer = 32, r_inner = 18) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  rad <- sqrt(cx^2 + cy^2)
  ifel(rad <= r_outer & rad >= r_inner, 1, 0)
}

make_L <- function(n = 81) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  values(r) <- 0
  r[10:70, 10:35] <- 1
  r[45:70, 10:70] <- 1
  r
}

make_patchy <- function(n = 81,
                                 num_patches = 5,
                                 fill = 0.15,
                                 buffer = TRUE) {
    
    stopifnot(
        num_patches > 0,
        num_patches < n * n,
        fill > 0,
        fill <= 1
    )
    
    target <- max(num_patches, floor(fill * n * n))
    
    ## raster
    mat <- matrix(0L, n, n)
    
    ## random seeds
    seeds <- sample.int(n * n, num_patches)
    mat[seeds] <- seq_len(num_patches)
    
    ## maximum possible frontier size
    max_frontier <- 2L * target
    
    frontier_cell <- integer(max_frontier)
    frontier_id   <- integer(max_frontier)
    
    frontier_cell[seq_len(num_patches)] <- seeds
    frontier_id[seq_len(num_patches)]   <- seq_len(num_patches)
    
    nf <- num_patches
    occupied <- num_patches
    
    ## 4-neighbour growth
    dr <- c(-1L, 1L, 0L, 0L)
    dc <- c(0L, 0L, -1L, 1L)
    
    while (occupied < target && nf > 0L) {
        
        ## random frontier element
        i <- sample.int(nf, 1L)
        
        cell <- frontier_cell[i]
        pid  <- frontier_id[i]
        
        ## remove by swap
        frontier_cell[i] <- frontier_cell[nf]
        frontier_id[i]   <- frontier_id[nf]
        nf <- nf - 1L
        
        r <- ((cell - 1L) %% n) + 1L
        c <- ((cell - 1L) %/% n) + 1L
        
        ord <- sample.int(4)
        
        for (k in ord) {
            
            rr <- r + dr[k]
            cc <- c + dc[k]
            
            if (rr < 1L || rr > n || cc < 1L || cc > n)
                next
            
            if (mat[rr, cc] != 0L)
                next
            
            if (buffer) {
                
                rs <- max(1L, rr - 1L):min(n, rr + 1L)
                cs <- max(1L, cc - 1L):min(n, cc + 1L)
                
                nbr <- mat[rs, cs]
                
                if (any(nbr > 0L & nbr != pid))
                    next
            }
            
            ## occupy
            mat[rr, cc] <- pid
            occupied <- occupied + 1L
            
            ## current cell remains active
            nf <- nf + 1L
            frontier_cell[nf] <- cell
            frontier_id[nf]   <- pid
            
            ## new cell becomes active
            nf <- nf + 1L
            frontier_cell[nf] <- (cc - 1L) * n + rr
            frontier_id[nf]   <- pid
            
            break
        }
    }
    
    mat[mat == 0L] <- NA_integer_
    mat[mat>0] <- 1 # Remove this if you want to preserve the patch id.
    
    rast(mat, crs="local")
}

shapes <- list(disk = make_disk(), pinwheel = make_pinwheel(), dumbbell = make_dumbbell(),
               ring = make_ring(), L = make_L(), patchy = make_patchy())

shapes_stack <- rast(shapes)
names(shapes_stack) <- names(shapes)
plot(shapes_stack, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, nc = 3)
```

# The fifteen indices

Each index has its own function. Naming convention follows `gm_*_index()` (e.g. `gm_moment_isotropy_index()`). Every index takes the
raster directly and returns a list with
`index` plus supporting detail (area, centroid, reference geometry for
plotting, and so on). `size` (number of random cells drawn) and `seed` (for reproducibility) are key inputs for the random draws. Called directly on a binary raster, they provide unweighted index:

```{r indices-direct}
gm_convexity_index(shapes$pinwheel, size = 2000, seed = 1)$index
gm_moment_isotropy_index(shapes$pinwheel)$index
gm_moment_isotropy_index(shapes$dumbbell)$index
gm_depth_index(shapes$ring)$index
gm_geodesic_span_index(shapes$dumbbell, size = 60, seed = 1)$index
```

`gm_shape_indices()` runs all of them (or a chosen subset via `which`) in
one call and returns a named vector. `size` here has to serve all five
Monte Carlo indices at once, including the two much more expensive
geodesic ones (see `?gm_shape_indices`), so it is preferable to keep it modest if computational time is a constraint.

```{r indices-combined}
all_results <- sapply(shapes, gm_shape_indices, size = 150, seed = 1)
knitr::kable(format = "html", round(all_results, 3))
```

A few things worth noticing in that table. **Pinwheel vs. dumbbell** is
the clearest illustration that `moment_isotropy` and `convexity` measure
genuinely different things: the four-bladed pinwheel has exact 4-fold
rotational symmetry, so its mass is distributed identically in every
direction - `moment_isotropy` scores essentially 1 - but it's deeply
non-convex (four spiky, empty notches), so `convexity`, `hull_ratio`, and
`reock` all score low. The **dumbbell** is a mirror image: elongated along
one axis, so `moment_isotropy` scores near 0, but a line between almost
any two interior points stays inside the shape (the connecting bar keeps
the two balls "in sight" of each other), so `convexity` stays
comparatively high. **The ring** scores lowest on `depth` of all six
shapes - a thin annulus has nowhere far from *some* boundary, inner or
outer - while its `hull_ratio` (0.67) reflects exactly what you'd expect:
the ring's own area over the *solid disk* that is its convex hull, since
punching a hole out of a shape never changes its hull. **The L-shape**
sits in between on most scores - visibly non-convex from its one notch,
but only in one place, not spiky like the pinwheel. **`patchy`** scores
sensibly on thirteen of the fifteen indices - lowest `hull_ratio` and
`polsby_popper` of any shape here, exactly as you'd expect from several
small, scattered pieces - but its `geodesic_span` and `geodesic_chord`
entries are `NA` (more on that below)

`gm_shape_indices()` also accepts a subset:

```{r indices-subset}
gm_shape_indices(shapes$dumbbell, which = c("hull_ratio", "reock", "detour"))
```

# Multi-part shapes

MultiPolygons are common place in vectors and an analogous situation are multipatch raters.
However a raster represents this with no special
machinery at all: every cell is independently inside or outside the
shape, so nothing about the representation itself distinguishes "one
piece" from "several disjoint piece, unless we explicitly determine the connectivity graph.

```{r patchy-patches}
p <- terra::patches(shapes$patchy, directions = 8, zeroAsNA = TRUE)
terra::global(p, "max", na.rm = TRUE)[[1]]  # number of disjoint pieces
```

Due to this representational convenience, many of the indices automatically deal with multiple parts and holes treating them equivalently (exterior is always background). However, this idea does NOT hold for `geodesic_span` and `geodesic_chord`:

```{r patchy-geodesic, warning = TRUE}
gm_geodesic_span_index(shapes$patchy, size = 60, seed = 1)$index
```

Geodesic distance is a path between two points, confined to the shape -
and there is no path between two points in different disjoint pieces, not
just a long one. 

It's tempting to reach for a workaround that still produces a number.
Two were checked directly and both make things
worse, not better:

- **Drop the unreachable point pairs and average over what's left.** For highly fragmented patch configuration, the reference disk stays the same as combined area, but filtered distances shrink towards whatever is typical *within* ever smaller patches and thus index grows beyond 1.

- **Score each patch separately and average the results.** identical patches close or far from one another will score the same. This is a legitimate, different metric in its own right, but better left for `landscapemetrics` and does not need to be duplicated here.

Thus, these functions return `NA` for multipatch rasters for these two indices.

## Computational cost: does the clean representation cost less?

Encoding multi-part shapes with no special machinery is a representation
question. Whether that also makes them cheaper to score is a separate,
empirical one.

```{r cost-check, include = FALSE}
eval_shapeindices <- requireNamespace("shapeindices", quietly = TRUE) &&
  requireNamespace("sf", quietly = TRUE)
```


```{r cost-scaling, eval = eval_shapeindices}
library(shapeindices)
library(sf)

raster_to_multipolygon <- function(r) sf::st_as_sf(terra::as.polygons(r, dissolve = TRUE))

scaling <- do.call(rbind, lapply(c(2, 5, 10, 20, 40, 80), function(np) {
  r <- make_patchy(n = 121, num_patches = np, fill = 0.15, buffer = TRUE)
  poly <- raster_to_multipolygon(r)
  t_gm  <- system.time(gm_shape_indices(r, size = 1500, seed = 1))
  t_vec <- system.time(shape_indices(poly, deterministic = FALSE, size = 1500, seed = 1))
  data.frame(patches = np, gridmorph = round(t_gm[["elapsed"]], 2),
             `shapeindices (Monte Carlo)` = round(t_vec[["elapsed"]], 2),
             check.names = FALSE)
}))
knitr::kable(format = "html", scaling, caption = "Seconds for all fifteen gridmorph indices vs. all thirteen shapeindices indices, same shapes, increasing patch count at roughly fixed total area.")
```

At fixed total area, `gridmorph`'s own cost stays essentially flat as
patch count increases - unsurprising, since valid-cell count barely
changes and every index here is `O(N)` or Monte Carlo `O(size)`,
regardless of how those cells are arranged into pieces (this holds for
`geodesic_span`/`geodesic_chord` too: the connectivity check above means
a disconnected shape returns its `NA` almost immediately, not after
paying for the sampling loop). `shapeindices`' own Monte Carlo mode grows
instead, because more patches means more separate boundary curves feeding
into the constrained Delaunay triangulation, and triangle count is what
its own cost scales with - not catastrophically here, but a real, visible
trend in the table above that `gridmorph`'s cell-count-based cost model
doesn't share.

```{r cost-deterministic, eval = eval_shapeindices}
deterministic <- do.call(rbind, lapply(c(2, 10, 40), function(np) {
  r <- make_patchy(n = 121, num_patches = np, fill = 0.15, buffer = TRUE)
  poly <- raster_to_multipolygon(r)
  n_tri <- nrow(cdt_triangles(poly))
  t <- system.time(shape_indices(poly, which = "convexity", deterministic = TRUE))
  data.frame(patches = np, n_triangles = n_tri, seconds = round(t[["elapsed"]], 2))
}))
knitr::kable(format = "html", deterministic, caption = "shapeindices' own deterministic (exhaustive) convexity_index() mode, same patchy shapes.")
```

That growth looks mild here because these patches are simple, well-
separated blobs, not realistically messy boundaries. `shapeindices`' own
deterministic mode is exhaustive over triangle PAIRS at a `n_quad^2`
multiplier per pair (`O(n_quad^2 * n^2)` - the package's own size-warning
message states this directly), and real boundaries produce far more
troublesome triangulations than tidy synthetic blobs do:
`vignette("c-comparison-with-shapeindices")`'s own NC-counties example
hits this directly on real county polygons - a single deterministic
`convexity_index()` call there takes on the order of 19 seconds at just
281 triangles, an order of magnitude worse than any patch count tested
here at a comparable triangle count. Multipolygons by itself is a mild
cost driver for `shapeindices`; boundary complexity is what actually
triggers the quadratic cliff, and messy real-world multi-part shapes tend
to have both at once.



# `weighted`: using the raster's own values as mass

For indices that admit weights, the weights are the values of the raster. For these indices `weigthed` defaults to `TRUE`, though can be reset by th user. The shape itself is always derived the same way - a cell counts as "inside" iff its own value is neither `NA` nor exactly `0`.  but `weighted` controls whether those *values* also act as a density/mass field throughout the computation, or are ignored beyond defining the shape. 

A plain binary mask (every inside cell equal to `1`) gives the identical answer either way, since there's no variation in mass to matter. A genuinely continuous raster is where the difference shows up, e.g.

```{r weighted-gradient, fig.width = 5, fig.height = 4}
gradient <- init(shapes$dumbbell, "x") + 41  # 0 (left) to ~120 (right)
weight_rast <- ifel(shapes$dumbbell == 1, gradient, NA)
plot(weight_rast, col = hcl.colors(50, "viridis"), axes = FALSE,
     main = "dumbbell, weighted by an x-gradient")
```

```{r weighted-compare}
plain    <- gm_moment_of_inertia_index(shapes$dumbbell)$index
ignored  <- gm_moment_of_inertia_index(weight_rast, weighted = FALSE)$index
weighted <- gm_moment_of_inertia_index(weight_rast, weighted = TRUE)$index
data.frame(plain, ignored, weighted)
```

`plain` and `ignored` match exactly - `weighted = FALSE` on a continuous
raster reproduces the plain binary-mask index precisely since it substitutes a constant mass over every valid cell regardless of the raster's own values. `weighted` differs: with more mass
concentrated toward the right ball, the shape's own effective centroid
shifts, changing the moment of inertia relative to the unweighted (geometric) case.

**One thing `weighted` does not change**: the six classic metrics
(`hull_ratio`, `polsby_popper`, `width_length_ratio`, `reock`, `detour`,
`exchange`) have no weighted form at all - they're properties of the
shape's own boundary, convex hull, or bounding circle alone, unaffected
by how mass is distributed inside it. 0-valued cells are considered outside the shape regardless.
They don't even have a `weighted` argument and when passed using `gm_shape_indices()`, it is silently ignored.

# Monte Carlo indices: `size`, `seed`

`gm_convexity_index()`, `gm_span_index()`, and
`gm_radial_concentration_index()` estimate their answer from a random
sample of points drawn from the raster's own valid cells - there's no
practical exhaustive alternative at raster-cell counts (see
`vignette("c-comparison-with-shapeindices")` for why). `size` controls how
many points are drawn (matching `terra::spatSample()`'s own argument
name); `seed` makes the draw reproducible:

```{r monte-carlo}
a <- gm_convexity_index(shapes$pinwheel, size = 500, seed = 42)$index
b <- gm_convexity_index(shapes$pinwheel, size = 500, seed = 42)$index
c <- gm_convexity_index(shapes$pinwheel, size = 500, seed = 43)$index
data.frame(seed_42_again = a, identical_to_first = identical(a, b), different_seed = c)
```

Larger `size` reduces sampling noise at the cost of more computation - a
few hundred is usually enough to see a shape's rough compactness; a few
thousand is closer to what you'd want for a final, reportable number.

However, while the two geodesic indices are Monte Carlo too, and take the same `size`
argument;`size = K` draws K
points and each one's own contribution is the exact mean of its own
`terra::gridDist()` field, so precision scales with `size` the same way
it does for the three indices above. What `size` does NOT unify is COST: each point here is a whole-raster `gridDist()` sweep, not a coordinate lookup, so it's a substantially more expensive index than the three above at the same `size`. This matters concretely for `gm_shape_indices()`: a `size` picked for should be picked for the most expensive requested index. 


```{r geodesic-monte-carlo}
gm_geodesic_span_index(shapes$pinwheel, size = 40, seed = 1)$index
gm_geodesic_span_index(shapes$pinwheel, size = 150, seed = 1)$index
```


# Morphological operators

`gridmorph` also exposes standard morphological operators directly on
`terra` rasters - `erode()`, `dilate()`, `opening()`, `closing()`,
`tophat()`, `bottomhat()` - something `terra` itself doesn't provide with
configurable structuring elements. A `kernel` argument takes any matrix
(`1` = included, `NA` = excluded), or one of three named shortcuts:

```{r kernels}
se_box(2)      # a 5x5 square
se_disc(2)     # a Euclidean disc of radius 2
se_diamond(2)  # a Manhattan (L1) diamond of radius 2
```

`opening()` (erode then dilate) removes small bright features narrower
than the structuring element - exactly the pinwheel's own thin,
tapering blade tips:

```{r morphology-demo, fig.width = 7, fig.height = 4}
pinwheel_opened <- opening(shapes$pinwheel, se_disc(3))
plot(c(shapes$pinwheel, pinwheel_opened), col = c("white", "steelblue"),
     legend = FALSE, axes = FALSE, main = c("pinwheel", "opening(pinwheel, se_disc(3))"))
```

```{r morphology-index-effect}
gm_convexity_index(shapes$pinwheel, size = 2000, seed = 1)$index
gm_convexity_index(pinwheel_opened, size = 2000, seed = 1)$index
```

Trimming the thinnest blade tips barely moves `convexity` here - the
pinwheel's own concavity comes from the gaps *between* blades, not the
tips themselves, so this particular structuring element isn't the tool
for that job. Morphological operators and shape indices answer different
questions: operators reshape a raster; indices score whatever shape you
hand them, opened or not.

None of the six operators require a 0/1 mask - they generalize to
continuous (grayscale) and categorical (label-image) rasters too, each in
a genuinely different way, not just a relabelled version of the binary
case. See `vignette("b-morphological-operators")` for the full tour.

# A note on coordinate systems

Every index in this package requires `rast` to use planar (projected)
coordinates - a geographic (longitude/latitude) CRS is rejected with an
error. This isn't because area, distance, or perimeter can't be computed
correctly on geographic data in general - `terra`'s own `expanse()`,
`perim()`, and `distance()` do that correctly, via geodesic math, even in
a geographic CRS. The real problem is narrower: this package's own
point-based geometry - line-crossing tests for `gm_convexity_index()`,
pairwise distances for `gm_span_index()`/`gm_radial_concentration_index()`,
moment-tensor calculations, minimum-enclosing-circle fitting - works
directly in `rast`'s own coordinate space as if it were Cartesian, and
isn't (yet) geodesic-aware. All six shapes above used `crs = "local"`, a
real (if arbitrary) projected CRS `terra` provides for exactly this kind
of synthetic-grid use - real-world rasters need a genuine projected CRS
(state plane, UTM, or similar) before calling any index here. 
