---
title: "1. Basic Usage"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
    code-fold: false
vignette: >
  %\VignetteIndexEntry{1. Basic Usage}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center")
```

```{r load}
library(gridmorph)
library(terra)
```

This vignette is a practical, code-first tour of the fifteen shape
indices: build a few synthetic raster shapes, run the indices on them
individually and all at once, and see what `weighted` changes.
`gridmorph` computes the same thirteen indices as its sibling package
`shapeindices` (which works on `sf` (multi)polygons instead of rasters) -
the underlying definitions and proofs are the same regardless of
representation, so they aren't repeated here - plus two indices with no
vector-package analogue at all (`gm_geodesic_span_index()` and
`gm_geodesic_chord_index()`, mean distance along the shortest path
CONFINED to the shape rather than a straight line - see the "Geodesic
indices" section below). For the mathematical grounding - what
"convexity" or "depth" even means, and where each index breaks down - see
`shapeindices`' own `vignette("b-understanding-convexity-index",
package = "shapeindices")` and its companions. For the six morphological
operators (`erode()`, `dilate()`, `opening()`, `closing()`, `tophat()`,
`bottomhat()`) and how they behave across binary, continuous, and
categorical rasters, see `vignette("b-morphological-operators")`. For how
`gridmorph` and `shapeindices` compare on accuracy, speed, and memory,
see `vignette("c-comparison-with-shapeindices")`. Six of the fifteen
indices compare the shape against a reference value computed the same
way the shape's own score was, rather than a plain textbook formula - see
`vignette("d-resolution-matched-references")` for why, and what a
perfectly rasterized disk scores on each as a result.

# Six canonical shapes

Six small synthetic rasters, built directly from cell coordinates, cover
a useful range of shape behaviour: a **disk** (the reference shape every
index treats as optimal), a **pinwheel** (four spiky, rotationally
symmetric blades), a **dumbbell** (two round masses joined by a thin
bar), a **ring** (a disk with its centre hollowed out), an **L-shape**
(a simple right-angle notch), and a **patchy** raster (several disjoint
patches grown independently, never touching each other - a genuine
multi-part shape, the raster analogue of an `sf` MULTIPOLYGON). The first
five are single, connected pieces; `patchy` is not, and shows up again in
its own section below.

```{r shapes}
#| code-fold: true
make_disk <- function(n = 81, radius = 30) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  ifel(sqrt(cx^2 + cy^2) <= radius, 1, 0)
}

make_pinwheel <- function(n = 81, blades = 4, r_max = 35) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  theta <- (atan2(cy, cx) + 2 * pi) %% (2 * pi)
  sector_width <- 2 * pi / blades
  theta_local <- theta %% sector_width
  # each blade sweeps from a point (theta_local = 0) out to r_max
  # (theta_local = sector_width) - `blades` copies, evenly rotated
  ifel(sqrt(cx^2 + cy^2) <= r_max * (theta_local / sector_width), 1, 0)
}

make_dumbbell <- function(n = 81, r_ball = 16, bar_half_width = 4, gap = 50) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  ball1 <- sqrt((cx + gap / 2)^2 + cy^2) <= r_ball
  ball2 <- sqrt((cx - gap / 2)^2 + cy^2) <= r_ball
  bar   <- (cx >= -gap / 2) & (cx <= gap / 2) & (abs(cy) <= bar_half_width)
  ifel(ball1 | ball2 | bar, 1, 0)
}

make_ring <- function(n = 81, r_outer = 32, r_inner = 18) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  rad <- sqrt(cx^2 + cy^2)
  ifel(rad <= r_outer & rad >= r_inner, 1, 0)
}

make_L <- function(n = 81) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  values(r) <- 0
  r[10:70, 10:35] <- 1
  r[45:70, 10:70] <- 1
  r
}

make_patchy <- function(n = 81,
                                 num_patches = 5,
                                 fill = 0.15,
                                 buffer = TRUE) {
    
    stopifnot(
        num_patches > 0,
        num_patches < n * n,
        fill > 0,
        fill <= 1
    )
    
    target <- max(num_patches, floor(fill * n * n))
    
    ## raster
    mat <- matrix(0L, n, n)
    
    ## random seeds
    seeds <- sample.int(n * n, num_patches)
    mat[seeds] <- seq_len(num_patches)
    
    ## maximum possible frontier size
    max_frontier <- 2L * target
    
    frontier_cell <- integer(max_frontier)
    frontier_id   <- integer(max_frontier)
    
    frontier_cell[seq_len(num_patches)] <- seeds
    frontier_id[seq_len(num_patches)]   <- seq_len(num_patches)
    
    nf <- num_patches
    occupied <- num_patches
    
    ## 4-neighbour growth
    dr <- c(-1L, 1L, 0L, 0L)
    dc <- c(0L, 0L, -1L, 1L)
    
    while (occupied < target && nf > 0L) {
        
        ## random frontier element
        i <- sample.int(nf, 1L)
        
        cell <- frontier_cell[i]
        pid  <- frontier_id[i]
        
        ## remove by swap
        frontier_cell[i] <- frontier_cell[nf]
        frontier_id[i]   <- frontier_id[nf]
        nf <- nf - 1L
        
        r <- ((cell - 1L) %% n) + 1L
        c <- ((cell - 1L) %/% n) + 1L
        
        ord <- sample.int(4)
        
        for (k in ord) {
            
            rr <- r + dr[k]
            cc <- c + dc[k]
            
            if (rr < 1L || rr > n || cc < 1L || cc > n)
                next
            
            if (mat[rr, cc] != 0L)
                next
            
            if (buffer) {
                
                rs <- max(1L, rr - 1L):min(n, rr + 1L)
                cs <- max(1L, cc - 1L):min(n, cc + 1L)
                
                nbr <- mat[rs, cs]
                
                if (any(nbr > 0L & nbr != pid))
                    next
            }
            
            ## occupy
            mat[rr, cc] <- pid
            occupied <- occupied + 1L
            
            ## current cell remains active
            nf <- nf + 1L
            frontier_cell[nf] <- cell
            frontier_id[nf]   <- pid
            
            ## new cell becomes active
            nf <- nf + 1L
            frontier_cell[nf] <- (cc - 1L) * n + rr
            frontier_id[nf]   <- pid
            
            break
        }
    }
    
    mat[mat == 0L] <- NA_integer_
    mat[mat>0] <- 1 # Remove this if you want to preserve the patch id.
    
    rast(mat, crs="local")
}

shapes <- list(disk = make_disk(), pinwheel = make_pinwheel(), dumbbell = make_dumbbell(),
               ring = make_ring(), L = make_L(), patchy = make_patchy())

shapes_stack <- rast(shapes)
names(shapes_stack) <- names(shapes)
plot(shapes_stack, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, nc = 3)
```

# The fifteen indices

Each index has its own function - `gm_depth_index()`,
`gm_moment_of_inertia_index()`, `gm_moment_isotropy_index()`,
`gm_directional_balance_index()`, `gm_convexity_index()`, `gm_span_index()`,
`gm_radial_concentration_index()` (Monte Carlo, drawn from the raster's own
valid cells), plus `gm_hull_ratio_index()`, `gm_polsby_popper_index()`,
`gm_width_length_ratio_index()`, `gm_reock_index()`, `gm_detour_index()`,
and `gm_exchange_index()` (classic redistricting-literature metrics: each
needs only the shape's own footprint, convex hull, or minimum bounding
circle) - the same thirteen `shapeindices` computes on a polygon. Two more
have no vector-package equivalent at all: `gm_geodesic_span_index()` and
`gm_geodesic_chord_index()` measure mean distance along the shortest path
CONFINED to the shape (`terra::gridDist()`), between two random interior
points or two random boundary points respectively, rather than a straight
line - the raster analogue of the "Traversal Index" (Angel, Parent &
Civco, 2010). They take the same `size`/`seed` as the three Monte Carlo
indices above, but cost far more per unit of `size` (see
`?gm_geodesic_span_index` for why), and, like the six classic metrics,
`gm_geodesic_chord_index()` has no `weighted` form. Every index takes the
raster directly - no separate mask argument - and returns a list with
`index` plus supporting detail (area, centroid, reference geometry for
plotting, and so on). Called directly on a shape:

```{r indices-direct}
gm_convexity_index(shapes$pinwheel, size = 2000, seed = 1)$index
gm_moment_isotropy_index(shapes$pinwheel)$index
gm_moment_isotropy_index(shapes$dumbbell)$index
gm_depth_index(shapes$ring)$index
gm_geodesic_span_index(shapes$dumbbell, size = 60, seed = 1)$index
```

Running all fifteen separately means writing fifteen calls.
`gm_shape_indices()` runs all of them (or a chosen subset via `which`) in
one call and returns a named vector. `size` here has to serve all five
Monte Carlo indices at once, including the two much more expensive
geodesic ones (see `?gm_shape_indices`), so it's kept modest rather than
reused from the `size = 2000` example above:

```{r indices-combined}
all_results <- sapply(shapes, gm_shape_indices, size = 150, seed = 1)
knitr::kable(format = "html", round(all_results, 3))
```

A few things worth noticing in that table. **Pinwheel vs. dumbbell** is
the clearest illustration that `moment_isotropy` and `convexity` measure
genuinely different things: the four-bladed pinwheel has exact 4-fold
rotational symmetry, so its mass is distributed identically in every
direction - `moment_isotropy` scores essentially 1 - but it's deeply
non-convex (four spiky, empty notches), so `convexity`, `hull_ratio`, and
`reock` all score low. The dumbbell is the mirror image: elongated along
one axis, so `moment_isotropy` scores near 0, but a line between almost
any two interior points stays inside the shape (the connecting bar keeps
the two balls "in sight" of each other), so `convexity` stays
comparatively high. **The ring** scores lowest on `depth` of all six
shapes - a thin annulus has nowhere far from *some* boundary, inner or
outer - while its `hull_ratio` (0.67) reflects exactly what you'd expect:
the ring's own area over the *solid disk* that is its convex hull, since
punching a hole out of a shape never changes its hull. **The L-shape**
sits in between on most scores - visibly non-convex from its one notch,
but only in one place, not spiky like the pinwheel. **`patchy`** scores
sensibly on thirteen of the fifteen indices - lowest `hull_ratio` and
`polsby_popper` of any shape here, exactly as you'd expect from several
small, scattered pieces - but its `geodesic_span` and `geodesic_chord`
entries are `NA`. The next section explains why, and what that does and
doesn't say about rasters and multi-part shapes.

`gm_shape_indices()` also accepts a subset:

```{r indices-subset}
gm_shape_indices(shapes$dumbbell, which = c("hull_ratio", "reock", "detour"))
```

# Multi-part shapes

`make_patchy()` (defined above, alongside the other five shape
generators) grows several patches independently from random seed cells
across the raster, refusing to let any two patches touch - the raster
equivalent of an `sf` MULTIPOLYGON with several disjoint parts, rather
than one connected piece. A raster represents this with no special
machinery at all: every cell is independently inside or outside the
shape, so nothing about the representation itself distinguishes "one
piece" from "several disjoint pieces." `terra::patches()` confirms
`patchy` really is multi-part, not just concave:

```{r patchy-patches}
p <- terra::patches(shapes$patchy, directions = 8, zeroAsNA = TRUE)
terra::global(p, "max", na.rm = TRUE)[[1]]  # number of disjoint pieces
```

That "no special machinery" claim holds for thirteen of the fifteen
indices - already visible in the results table above, where `patchy` got
an ordinary numeric score on every area/moment/hull/boundary-based index,
computed exactly the same way as for the five single-piece shapes. It
does NOT hold for `geodesic_span` and `geodesic_chord`:

```{r patchy-geodesic, warning = TRUE}
gm_geodesic_span_index(shapes$patchy, size = 60, seed = 1)$index
```

Geodesic distance is a path between two points, confined to the shape -
and there is no path between two points in different disjoint pieces, not
just a long one. That's not a computational shortfall to work around; it
is what "disconnected" means. `NA` with an explicit warning is the
correct answer, not a gap in this package's own implementation - the
warning fires from a cheap up-front connectivity check
(`terra::patches()` again, internally), not from wasting time sampling
first and discovering the same fact afterward.

It's tempting to reach for a workaround that still produces a number.
Two were checked directly, not just argued about, and both make things
worse, not better:

- **Drop the unreachable point pairs and average over what's left.**
  Tested on shapes holding total area roughly fixed while patch count
  increased: the resulting "index" climbed from 1.05 at one patch to
  4.96 at 25 patches - a badly FRAGMENTED shape scoring nearly five times
  MORE compact than a disk, because the reference disk stays sized to the
  full combined area while the filtered distances shrink toward
  whatever's typical *within* an ever-smaller patch. The more fragmented
  the shape, the more misleading the number.
- **Score each patch separately and average the results.** Tested on
  four identical small patches, once placed close together and once
  scattered across the same raster: both arrangements scored identically
  (1.001 either way), because per-patch averaging throws away everything
  about how the patches relate to each other spatially - exactly the
  information "treat the whole mask as one shape" exists to capture.
  (This is a legitimate, different metric in its own right - it's what
  `landscapemetrics`' class-level functions already compute, not
  something gridmorph needs to duplicate.)

`NA` isn't a limitation to route around; it's the honest answer for a
quantity that genuinely has none, for these two indices only.

## Computational cost: does the clean representation cost less?

Encoding multi-part shapes with no special machinery is a representation
question. Whether that also makes them cheaper to score is a separate,
empirical one.

```{r cost-check, include = FALSE}
eval_shapeindices <- requireNamespace("shapeindices", quietly = TRUE) &&
  requireNamespace("sf", quietly = TRUE)
```

```{r cost-not-available, eval = !eval_shapeindices, echo = FALSE, results = "asis"}
cat("`shapeindices` isn't installed here, so the cost comparison below ",
    "doesn't run. Install `shapeindices` and `sf` to reproduce it.")
```

```{r cost-scaling, eval = eval_shapeindices}
library(shapeindices)
library(sf)

raster_to_multipolygon <- function(r) sf::st_as_sf(terra::as.polygons(r, dissolve = TRUE))

scaling <- do.call(rbind, lapply(c(2, 5, 10, 20, 40, 80), function(np) {
  r <- make_patchy(n = 121, num_patches = np, fill = 0.15, buffer = TRUE)
  poly <- raster_to_multipolygon(r)
  t_gm  <- system.time(gm_shape_indices(r, size = 1500, seed = 1))
  t_vec <- system.time(shape_indices(poly, deterministic = FALSE, size = 1500, seed = 1))
  data.frame(patches = np, gridmorph = round(t_gm[["elapsed"]], 2),
             `shapeindices (Monte Carlo)` = round(t_vec[["elapsed"]], 2),
             check.names = FALSE)
}))
knitr::kable(format = "html", scaling, caption = "Seconds for all fifteen gridmorph indices vs. all thirteen shapeindices indices, same shapes, increasing patch count at roughly fixed total area.")
```

At fixed total area, `gridmorph`'s own cost stays essentially flat as
patch count increases - unsurprising, since valid-cell count barely
changes and every index here is `O(N)` or Monte Carlo `O(size)`,
regardless of how those cells are arranged into pieces (this holds for
`geodesic_span`/`geodesic_chord` too: the connectivity check above means
a disconnected shape returns its `NA` almost immediately, not after
paying for the sampling loop). `shapeindices`' own Monte Carlo mode grows
instead, because more patches means more separate boundary curves feeding
into the constrained Delaunay triangulation, and triangle count is what
its own cost scales with - not catastrophically here, but a real, visible
trend in the table above that `gridmorph`'s cell-count-based cost model
doesn't share.

```{r cost-deterministic, eval = eval_shapeindices}
deterministic <- do.call(rbind, lapply(c(2, 10, 40), function(np) {
  r <- make_patchy(n = 121, num_patches = np, fill = 0.15, buffer = TRUE)
  poly <- raster_to_multipolygon(r)
  n_tri <- nrow(cdt_triangles(poly))
  t <- system.time(shape_indices(poly, which = "convexity", deterministic = TRUE))
  data.frame(patches = np, n_triangles = n_tri, seconds = round(t[["elapsed"]], 2))
}))
knitr::kable(format = "html", deterministic, caption = "shapeindices' own deterministic (exhaustive) convexity_index() mode, same patchy shapes.")
```

That growth looks mild here because these patches are simple, well-
separated blobs, not realistically messy boundaries. `shapeindices`' own
deterministic mode is exhaustive over triangle PAIRS at a `n_quad^2`
multiplier per pair (`O(n_quad^2 * n^2)` - the package's own size-warning
message states this directly), and real boundaries produce far more
troublesome triangulations than tidy synthetic blobs do:
`vignette("c-comparison-with-shapeindices")`'s own NC-counties example
hits this directly on real county polygons - a single deterministic
`convexity_index()` call there takes on the order of 19 seconds at just
281 triangles, an order of magnitude worse than any patch count tested
here at a comparable triangle count. Patch count by itself is a mild
cost driver for `shapeindices`; boundary complexity is what actually
triggers the quadratic cliff, and messy real-world multi-part shapes tend
to have both at once.

So: the representation question and the cost question have different
answers. Encoding a multi-part shape needs zero special-casing on a
raster (thirteen of fifteen indices) and a structural, not incidental,
limitation on the other two (geodesic distance has no value across
disconnected pieces, by definition, not by implementation gap). Cost is
a separate story - `gridmorph` stays flat with patch count because its
cost model was never coupled to boundary complexity in the first place,
while `shapeindices` grows because its mesh does, mildly for simple
patches and severely for realistically complex ones. Cleaner encoding and
lower cost aren't the same claim, and only one of them holds
unconditionally here.

# `weighted`: using the raster's own values as mass

Every index function takes `weighted` (default `TRUE`). The shape itself
is always derived the same way - a cell counts as "inside" iff its own
value is neither `NA` nor exactly `0` - but `weighted` controls whether
those *values* also act as a density/mass field throughout the
computation, or are ignored beyond defining the shape. A plain binary
mask (every inside cell equal to `1`) gives the identical answer either
way, since there's no variation in mass to matter. A genuinely continuous
raster is where the difference shows up - here, a gradient increasing
toward the dumbbell's right-hand ball:

```{r weighted-gradient, fig.width = 5, fig.height = 4}
gradient <- init(shapes$dumbbell, "x") + 41  # 0 (left) to ~120 (right)
weight_rast <- ifel(shapes$dumbbell == 1, gradient, NA)
plot(weight_rast, col = hcl.colors(50, "viridis"), axes = FALSE,
     main = "dumbbell, weighted by an x-gradient")
```

```{r weighted-compare}
plain    <- gm_moment_of_inertia_index(shapes$dumbbell)$index
ignored  <- gm_moment_of_inertia_index(weight_rast, weighted = FALSE)$index
weighted <- gm_moment_of_inertia_index(weight_rast, weighted = TRUE)$index
data.frame(plain, ignored, weighted)
```

`plain` and `ignored` match exactly - `weighted = FALSE` on a continuous
raster reproduces the plain binary-mask index precisely, not
approximately, since it substitutes a constant mass over every valid cell
regardless of the raster's own values. `weighted` differs: with more mass
concentrated toward the right ball, the shape's own effective centroid
shifts, changing the moment of inertia relative to the unweighted
(geometric) case.

**One thing `weighted` does not change**: the six classic metrics
(`hull_ratio`, `polsby_popper`, `width_length_ratio`, `reock`, `detour`,
`exchange`) have no weighted form at all - they're properties of the
shape's own boundary, convex hull, or bounding circle alone, unaffected
by how mass is distributed inside it. 0-valued cells are considered outside the shape regardless.
They don't even have a `weighted` argument and when passed using `gm_shape_indices()`, it is silently ignored.

# Monte Carlo indices: `size`, `seed`

`gm_convexity_index()`, `gm_span_index()`, and
`gm_radial_concentration_index()` estimate their answer from a random
sample of points drawn from the raster's own valid cells - there's no
practical exhaustive alternative at raster-cell counts (see
`vignette("c-comparison-with-shapeindices")` for why). `size` controls how
many points are drawn (matching `terra::spatSample()`'s own argument
name); `seed` makes the draw reproducible:

```{r monte-carlo}
a <- gm_convexity_index(shapes$pinwheel, size = 500, seed = 42)$index
b <- gm_convexity_index(shapes$pinwheel, size = 500, seed = 42)$index
c <- gm_convexity_index(shapes$pinwheel, size = 500, seed = 43)$index
data.frame(seed_42_again = a, identical_to_first = identical(a, b), different_seed = c)
```

Larger `size` reduces sampling noise at the cost of more computation - a
few hundred is usually enough to see a shape's rough compactness; a few
thousand is closer to what you'd want for a final, reportable number.

## `gm_geodesic_span_index()`/`gm_geodesic_chord_index()`: same `size`, very different cost

The two geodesic indices are Monte Carlo too, and take the same `size`
argument, meaning the same thing statistically: `size = K` draws K
points and each one's own contribution is the exact mean of its own
`terra::gridDist()` field, so precision scales with `size` the same way
it does for the three indices above (`?gm_geodesic_span_index` has the
full reasoning, including why an earlier version of this package used a
separate `n_points` argument here and no longer does).

What `size` does NOT unify is COST: each point here is a whole-raster
`gridDist()` sweep, not a coordinate lookup, so it's a substantially more
expensive index than the three above at the same `size`. This matters
concretely for `gm_shape_indices()`: a `size` picked for those three
(a few hundred to a few thousand) can be far too slow, or exceed this
index's own much stricter memory/time ceiling, once it's also reached
through `gm_shape_indices()`'s shared `...` - pick `size` with the most
expensive requested index in mind, not just the cheapest.

```{r geodesic-monte-carlo}
gm_geodesic_span_index(shapes$pinwheel, size = 40, seed = 1)$index
gm_geodesic_span_index(shapes$pinwheel, size = 150, seed = 1)$index
```

Even so, `D` and `D_ref` carry more Monte Carlo noise here than the
closed-form-reference indices above do at a comparable `size`:
`gridDist()`'s own angular quantization on a discrete grid adds
variability on top of ordinary point-sampling noise, so getting a
comparably precise answer needs a larger `size` than the three indices
above would.

# Morphological operators

`gridmorph` also exposes standard morphological operators directly on
`terra` rasters - `erode()`, `dilate()`, `opening()`, `closing()`,
`tophat()`, `bottomhat()` - something `terra` itself doesn't provide with
configurable structuring elements. A `kernel` argument takes any matrix
(`1` = included, `NA` = excluded), or one of three named shortcuts:

```{r kernels}
se_box(2)      # a 5x5 square
se_disc(2)     # a Euclidean disc of radius 2
se_diamond(2)  # a Manhattan (L1) diamond of radius 2
```

`opening()` (erode then dilate) removes small bright features narrower
than the structuring element - exactly the pinwheel's own thin,
tapering blade tips:

```{r morphology-demo, fig.width = 7, fig.height = 4}
pinwheel_opened <- opening(shapes$pinwheel, se_disc(3))
plot(c(shapes$pinwheel, pinwheel_opened), col = c("white", "steelblue"),
     legend = FALSE, axes = FALSE, main = c("pinwheel", "opening(pinwheel, se_disc(3))"))
```

```{r morphology-index-effect}
gm_convexity_index(shapes$pinwheel, size = 2000, seed = 1)$index
gm_convexity_index(pinwheel_opened, size = 2000, seed = 1)$index
```

Trimming the thinnest blade tips barely moves `convexity` here - the
pinwheel's own concavity comes from the gaps *between* blades, not the
tips themselves, so this particular structuring element isn't the tool
for that job. Morphological operators and shape indices answer different
questions: operators reshape a raster; indices score whatever shape you
hand them, opened or not.

None of the six operators require a 0/1 mask - they generalize to
continuous (grayscale) and categorical (label-image) rasters too, each in
a genuinely different way, not just a relabelled version of the binary
case. See `vignette("b-morphological-operators")` for the full tour,
including why `tophat()`/`bottomhat()` return a real-valued residual on
continuous input but a boolean flag on categorical input.

# A note on coordinate systems

Every index in this package requires `rast` to use planar (projected)
coordinates - a geographic (longitude/latitude) CRS is rejected with an
error. This isn't because area, distance, or perimeter can't be computed
correctly on geographic data in general - `terra`'s own `expanse()`,
`perim()`, and `distance()` do that correctly, via geodesic math, even in
a geographic CRS. The real problem is narrower: this package's own
point-based geometry - line-crossing tests for `gm_convexity_index()`,
pairwise distances for `gm_span_index()`/`gm_radial_concentration_index()`,
moment-tensor calculations, minimum-enclosing-circle fitting - works
directly in `rast`'s own x/y coordinate space as if it were Cartesian, and
isn't yet geodesic-aware. All six shapes above used `crs = "local"`, a
real (if arbitrary) projected CRS `terra` provides for exactly this kind
of synthetic-grid use - real-world rasters need a genuine projected CRS
(state plane, UTM, or similar) before calling any index here. See
`?gridmorph` for the full policy, including what happens with no CRS set
at all.
