---
title: "1. Basic Usage"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
vignette: >
  %\VignetteIndexEntry{1. Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center")
```

```{r load}
library(gridmorph)
library(terra)
```

This vignette is a practical, code-first tour of the thirteen shape
indices: build a few synthetic raster shapes, run the indices on them
individually and all at once, and see what `weighted` changes.
`gridmorph` computes the same thirteen indices as its sibling package
`shapeindices` (which works on `sf` (multi)polygons instead of rasters) -
the underlying definitions and proofs are the same regardless of
representation, so they aren't repeated here. For the mathematical
grounding - what "convexity" or "depth" even means, and where each index
breaks down - see `shapeindices`' own `vignette("b-understanding-convexity-index",
package = "shapeindices")` and its companions. For the six morphological
operators (`erode()`, `dilate()`, `opening()`, `closing()`, `tophat()`,
`bottomhat()`) and how they behave across binary, continuous, and
categorical rasters, see `vignette("b-morphological-operators")`. For how
`gridmorph` and `shapeindices` compare on accuracy, speed, and memory,
see `vignette("c-comparison-with-shapeindices")`. Four of the thirteen
indices compare the shape against a reference value computed the same
way the shape's own score was, rather than a plain textbook formula - see
`vignette("d-resolution-matched-references")` for why, and what a
perfectly rasterized disk scores on each as a result.

# Five canonical shapes

Five small synthetic rasters, built directly from cell coordinates, cover
a useful range of shape behaviour: a **disk** (the reference shape every
index treats as optimal), a **pinwheel** (four spiky, rotationally
symmetric blades), a **dumbbell** (two round masses joined by a thin
bar), a **ring** (a disk with its centre hollowed out), and an **L-shape**
(a simple right-angle notch).

```{r shapes}
make_disk <- function(n = 81, radius = 30) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  ifel(sqrt(cx^2 + cy^2) <= radius, 1, 0)
}

make_pinwheel <- function(n = 81, blades = 4, r_max = 35) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  theta <- (atan2(cy, cx) + 2 * pi) %% (2 * pi)
  sector_width <- 2 * pi / blades
  theta_local <- theta %% sector_width
  # each blade sweeps from a point (theta_local = 0) out to r_max
  # (theta_local = sector_width) - `blades` copies, evenly rotated
  ifel(sqrt(cx^2 + cy^2) <= r_max * (theta_local / sector_width), 1, 0)
}

make_dumbbell <- function(n = 81, r_ball = 16, bar_half_width = 4, gap = 50) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  ball1 <- sqrt((cx + gap / 2)^2 + cy^2) <= r_ball
  ball2 <- sqrt((cx - gap / 2)^2 + cy^2) <= r_ball
  bar   <- (cx >= -gap / 2) & (cx <= gap / 2) & (abs(cy) <= bar_half_width)
  ifel(ball1 | ball2 | bar, 1, 0)
}

make_ring <- function(n = 81, r_outer = 32, r_inner = 18) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  rad <- sqrt(cx^2 + cy^2)
  ifel(rad <= r_outer & rad >= r_inner, 1, 0)
}

make_L <- function(n = 81) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  values(r) <- 0
  r[10:70, 10:35] <- 1
  r[45:70, 10:70] <- 1
  r
}

shapes <- list(disk = make_disk(), pinwheel = make_pinwheel(), dumbbell = make_dumbbell(),
               ring = make_ring(), L = make_L())

shapes_stack <- rast(shapes)
names(shapes_stack) <- names(shapes)
plot(shapes_stack, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, nc = 3)
```

# The thirteen indices

Each index has its own function - `gm_depth_index()`,
`gm_moment_of_inertia_index()`, `gm_moment_isotropy_index()`,
`gm_directional_balance_index()`, `gm_convexity_index()`, `gm_span_index()`,
`gm_radial_concentration_index()` (Monte Carlo, drawn from the raster's own
valid cells), plus `gm_hull_ratio_index()`, `gm_polsby_popper_index()`,
`gm_width_length_ratio_index()`, `gm_reock_index()`, `gm_detour_index()`,
and `gm_exchange_index()` (classic redistricting-literature metrics: each
needs only the shape's own footprint, convex hull, or minimum bounding
circle). Every one takes the raster directly - no separate mask argument -
and returns a list with `index` plus supporting detail (area, centroid,
reference geometry for plotting, and so on). Called directly on a shape:

```{r indices-direct}
gm_convexity_index(shapes$pinwheel, size = 2000, seed = 1)$index
gm_moment_isotropy_index(shapes$pinwheel)$index
gm_moment_isotropy_index(shapes$dumbbell)$index
gm_depth_index(shapes$ring)$index
```

Running all thirteen separately means writing thirteen calls.
`gm_shape_indices()` runs all of them (or a chosen subset via `which`) in
one call and returns a named vector:

```{r indices-combined}
all_results <- sapply(shapes, gm_shape_indices, size = 2000, seed = 1)
knitr::kable(round(all_results, 3))
```

A few things worth noticing in that table. **Pinwheel vs. dumbbell** is
the clearest illustration that `moment_isotropy` and `convexity` measure
genuinely different things: the four-bladed pinwheel has exact 4-fold
rotational symmetry, so its mass is distributed identically in every
direction - `moment_isotropy` scores essentially 1 - but it's deeply
non-convex (four spiky, empty notches), so `convexity`, `hull_ratio`, and
`reock` all score low. The dumbbell is the mirror image: elongated along
one axis, so `moment_isotropy` scores near 0, but a line between almost
any two interior points stays inside the shape (the connecting bar keeps
the two balls "in sight" of each other), so `convexity` stays
comparatively high. **The ring** scores lowest on `depth` of all five
shapes - a thin annulus has nowhere far from *some* boundary, inner or
outer - while its `hull_ratio` (0.67) reflects exactly what you'd expect:
the ring's own area over the *solid disk* that is its convex hull, since
punching a hole out of a shape never changes its hull. **The L-shape**
sits in between on most scores - visibly non-convex from its one notch,
but only in one place, not spiky like the pinwheel.

`gm_shape_indices()` also accepts a subset:

```{r indices-subset}
gm_shape_indices(shapes$dumbbell, which = c("hull_ratio", "reock", "detour"))
```

# `weighted`: using the raster's own values as mass

Every index function takes `weighted` (default `TRUE`). The shape itself
is always derived the same way - a cell counts as "inside" iff its own
value is neither `NA` nor exactly `0` - but `weighted` controls whether
those *values* also act as a density/mass field throughout the
computation, or are ignored beyond defining the shape. A plain binary
mask (every inside cell equal to `1`) gives the identical answer either
way, since there's no variation in mass to matter. A genuinely continuous
raster is where the difference shows up - here, a gradient increasing
toward the dumbbell's right-hand ball:

```{r weighted-gradient, fig.width = 5, fig.height = 4}
gradient <- init(shapes$dumbbell, "x") + 41  # 0 (left) to ~120 (right)
weight_rast <- ifel(shapes$dumbbell == 1, gradient, NA)
plot(weight_rast, col = hcl.colors(50, "viridis"), axes = FALSE,
     main = "dumbbell, weighted by an x-gradient")
```

```{r weighted-compare}
plain    <- gm_moment_of_inertia_index(shapes$dumbbell)$index
ignored  <- gm_moment_of_inertia_index(weight_rast, weighted = FALSE)$index
weighted <- gm_moment_of_inertia_index(weight_rast, weighted = TRUE)$index
data.frame(plain, ignored, weighted)
```

`plain` and `ignored` match exactly - `weighted = FALSE` on a continuous
raster reproduces the plain binary-mask index precisely, not
approximately, since it substitutes a constant mass over every valid cell
regardless of the raster's own values. `weighted` differs: with more mass
concentrated toward the right ball, the shape's own effective centroid
shifts, changing the moment of inertia relative to the unweighted
(geometric) case.

**One thing `weighted` does not change**: the six classic metrics
(`hull_ratio`, `polsby_popper`, `width_length_ratio`, `reock`, `detour`,
`exchange`) have no weighted form at all - they're properties of the
shape's own boundary, convex hull, or bounding circle alone, unaffected
by how mass is distributed inside it. 0-valued cells are considered outside the shape regardless.
They don't even have a `weighted` argument and when passed using `gm_shape_indices()`, it is silently ignored.

# Monte Carlo indices: `size`, `seed`

`gm_convexity_index()`, `gm_span_index()`, and
`gm_radial_concentration_index()` estimate their answer from a random
sample of points drawn from the raster's own valid cells - there's no
practical exhaustive alternative at raster-cell counts (see
`vignette("c-comparison-with-shapeindices")` for why). `size` controls how
many points are drawn (matching `terra::spatSample()`'s own argument
name); `seed` makes the draw reproducible:

```{r monte-carlo}
a <- gm_convexity_index(shapes$pinwheel, size = 500, seed = 42)$index
b <- gm_convexity_index(shapes$pinwheel, size = 500, seed = 42)$index
c <- gm_convexity_index(shapes$pinwheel, size = 500, seed = 43)$index
data.frame(seed_42_again = a, identical_to_first = identical(a, b), different_seed = c)
```

Larger `size` reduces sampling noise at the cost of more computation - a
few hundred is usually enough to see a shape's rough compactness; a few
thousand is closer to what you'd want for a final, reportable number.

# Morphological operators

`gridmorph` also exposes standard morphological operators directly on
`terra` rasters - `erode()`, `dilate()`, `opening()`, `closing()`,
`tophat()`, `bottomhat()` - something `terra` itself doesn't provide with
configurable structuring elements. A `kernel` argument takes any matrix
(`1` = included, `NA` = excluded), or one of three named shortcuts:

```{r kernels}
se_box(2)      # a 5x5 square
se_disc(2)     # a Euclidean disc of radius 2
se_diamond(2)  # a Manhattan (L1) diamond of radius 2
```

`opening()` (erode then dilate) removes small bright features narrower
than the structuring element - exactly the pinwheel's own thin,
tapering blade tips:

```{r morphology-demo, fig.width = 7, fig.height = 4}
pinwheel_opened <- opening(shapes$pinwheel, se_disc(3))
plot(c(shapes$pinwheel, pinwheel_opened), col = c("white", "steelblue"),
     legend = FALSE, axes = FALSE, main = c("pinwheel", "opening(pinwheel, se_disc(3))"))
```

```{r morphology-index-effect}
gm_convexity_index(shapes$pinwheel, size = 2000, seed = 1)$index
gm_convexity_index(pinwheel_opened, size = 2000, seed = 1)$index
```

Trimming the thinnest blade tips barely moves `convexity` here - the
pinwheel's own concavity comes from the gaps *between* blades, not the
tips themselves, so this particular structuring element isn't the tool
for that job. Morphological operators and shape indices answer different
questions: operators reshape a raster; indices score whatever shape you
hand them, opened or not.

None of the six operators require a 0/1 mask - they generalize to
continuous (grayscale) and categorical (label-image) rasters too, each in
a genuinely different way, not just a relabelled version of the binary
case. See `vignette("b-morphological-operators")` for the full tour,
including why `tophat()`/`bottomhat()` return a real-valued residual on
continuous input but a boolean flag on categorical input.

# A note on coordinate systems

Every index in this package requires `rast` to use planar (projected)
coordinates - a geographic (longitude/latitude) CRS is rejected with an
error. This isn't because area, distance, or perimeter can't be computed
correctly on geographic data in general - `terra`'s own `expanse()`,
`perim()`, and `distance()` do that correctly, via geodesic math, even in
a geographic CRS. The real problem is narrower: this package's own
point-based geometry - line-crossing tests for `gm_convexity_index()`,
pairwise distances for `gm_span_index()`/`gm_radial_concentration_index()`,
moment-tensor calculations, minimum-enclosing-circle fitting - works
directly in `rast`'s own x/y coordinate space as if it were Cartesian, and
isn't yet geodesic-aware. All five shapes above used `crs = "local"`, a
real (if arbitrary) projected CRS `terra` provides for exactly this kind
of synthetic-grid use - real-world rasters need a genuine projected CRS
(state plane, UTM, or similar) before calling any index here. See
`?gridmorph` for the full policy, including what happens with no CRS set
at all.
