---
title: "2. Morphological Operators"
author: Nikhil Kaza
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
vignette: >
  %\VignetteIndexEntry{2. Morphological Operators}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center")
```

```{r load}
library(gridmorph)
library(terra)
```

`gridmorph` exposes six standard morphological operators - `erode()`,
`dilate()`, `opening()`, `closing()`, `tophat()`, `bottomhat()` - directly
on `terra` SpatRaster objects, something `terra` itself doesn't provide
with configurable structuring elements. This vignette is a full tour of
them, focused specifically on the part that's easy to get wrong: **what
each operator actually means changes depending on whether the input
raster is binary, continuous, or categorical**, and that's not just a
detail - two of the six operators (`tophat()`, `bottomhat()`) compute a
genuinely different kind of result for categorical input than for the
other two. 

# Structuring elements

Every operator takes a `kernel` argument: a plain matrix where `1`
includes a cell in the structuring element's footprint and `NA` excludes
it (matching `terra::focal()`'s own `w` convention directly). Three named
shortcuts build the common shapes:

```{r kernels}
se_box(2)      # a 5x5 square
se_disc(2)     # a Euclidean disc of radius 2
```

Any hand-built matrix works too, as long as its non-`NA` entries are all
exactly `1`. The operators reject anything else with a clear
error rather than silently misbehaving.

# Binary masks

This is the operators' original, still fully supported case: `1` =
inside the shape, `0`/`NA` = outside. `erode()` shrinks a shape by one
structuring-element radius; `dilate()` grows it:

```{r binary-erode-dilate, fig.width = 7, fig.height = 4}
r <- rast(nrows = 61, ncols = 61, xmin = 0, xmax = 61, ymin = 0, ymax = 61, crs = "local")
cx <- init(r, "x") - 30.5
cy <- init(r, "y") - 30.5
disk <- ifel(sqrt(cx^2 + cy^2) <= 22, 1, 0)

par(mfrow = c(1, 3))
plot(erode(disk, se_disc(4)), col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "erode")
plot(disk, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "original")
plot(dilate(disk, se_disc(4)), col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "dilate")
```

`opening()` (erode then dilate) removes small bright features narrower
than the structuring element without shrinking the shape overall;
`closing()` (dilate then erode) fills small dark gaps without growing it.
Both compose directly - `opening(mask, k)` is exactly
`dilate(erode(mask, k), k)`, not an independent implementation - which
makes them a natural denoising pair for classification noise: random
single-cell flips (both spurious "on" cells and spurious "off" holes)
mostly disappear after `opening()` then `closing()`:

```{r binary-denoise, fig.width = 7, fig.height = 4}
set.seed(1)
noisy <- disk
flipped <- sample(ncell(noisy), 80)
values(noisy)[flipped] <- 1 - values(noisy)[flipped]

cleaned1 <- closing(opening(noisy, se_disc(1)), se_disc(1))
cleaned2 <- opening(closing(noisy, se_disc(1)), se_disc(1))

par(mfrow = c(1, 3))
plot(noisy, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "noisy (80 flipped cells)")
plot(cleaned1, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "opening then closing")
plot(cleaned2, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "closing then opening")

cat("cells differing from the true disk \n",
     "noisy:", sum(values(noisy) != values(disk)), "\n",
     "O-C:", sum(values(cleaned1) != values(disk)), "\n",
     "C-O:", sum(values(cleaned2) != values(disk)))
```

Note however, that these operators are not commutative.

`tophat()` (`k AND NOT opening(k)`) highlights exactly the small
bright features `opening()` removed; `bottomhat()`
(`closing(k) AND NOT k`) highlights exactly the small dark gaps
`closing()` filled. Both return a `0`/`1` raster, same grid as the input.

```{r hat_demo, fig.width = 7, fig.height = 4}
h1 <- tophat(noisy, se_box(1))
h2 <- bottomhat(noisy, se_box(1))

par(mfrow = c(1, 3))
plot(noisy, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "noisy (80 flipped cells)")
plot(h1, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "Tophat")
plot(h2, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "Bottomhat")
```

# Continuous rasters

None of the six operators require a 0/1 binary raster `erode()`/`dilate()` are
`focal()`-based local-minimum/local-maximum filters with no binarization
step, so on continuous input they compute ordinary *grayscale/continuous*
morphology - a standard generalization of the binary case. `opening()`/`closing()`/`tophat()`/`bottomhat()`
inherit that directly, since each is built from `erode()`/`dilate()`
alone. On a smooth left-to-right gradient, `erode()` visibly pulls the
whole surface down toward the background outside the shape (a local
*minimum* filter), and `dilate()` pulls it up (a local *maximum* filter):

```{r continuous-erode-dilate, fig.width = 10, fig.height = 4}
disk_r <- rast(nrows = 51, ncols = 51, xmin = 0, xmax = 51, ymin = 0, ymax = 51, crs = "local")
gcx <- init(disk_r, "x") - 25.5
gcy <- init(disk_r, "y") - 25.5
disk_mask <- ifel(sqrt(gcx^2 + gcy^2) <= 20, 1, 0)
gradient <- ifel(disk_mask == 1, gcx + 30, NA)
eroded <- erode(gradient, se_disc(4))
dilated <- dilate(gradient, se_disc(4))
shared_range <- range(c(values(eroded), values(gradient), values(dilated)), na.rm = TRUE)

par(mfrow = c(1, 3))
plot(gradient, col = hcl.colors(50, "viridis"), range = shared_range, axes = FALSE, main = "gradient")
plot(eroded, col = hcl.colors(50, "viridis"), range = shared_range, axes = FALSE, main = "erode(gradient, se_disc(4))")
plot(dilated, col = hcl.colors(50, "viridis"), range = shared_range, axes = FALSE, main = "dilate(gradient, se_disc(4))")
```

Note that `erode()`/`dilate()` fill the WHOLE raster with a real, defined
value (the dark background outside the shrunk/grown shape is `0`, not
`NA`) - they reshape the raster over its entire extent, the same way
`erode()`/`dilate()` always have for a plain binary mask.
`tophat()`/`bottomhat()` are different: their job is to flag a specific
*feature*, so their output stays `NA` whenever `mask` itself was `NA`,
rather than asserting a defined `0` residual for a location that was
never part of the shape to begin with. A flat plateau with one small,
deliberate bump makes the real-valued residual unambiguous - `opening()`
(with a structuring element bigger than the bump) erases it completely,
so `tophat()` flags exactly the bump and nothing else:

```{r continuous-tophat, fig.width = 10, fig.height = 4}
plateau <- ifel(disk_mask == 1, 50, NA)
bump <- ifel(sqrt((gcx - 10)^2 + (gcy + 5)^2) <= 2, 20, 0)  # radius 2, smaller than the se_disc(3) below
weight_rast <- plateau + bump

weight_th <- tophat(weight_rast, se_disc(3))
par(mfrow = c(1, 2))
plot(weight_rast, col = hcl.colors(50, "viridis"), axes = FALSE, main = "weight_rast (plateau + one bump)")
plot(weight_th, col = hcl.colors(50, "viridis"), axes = FALSE, main = "tophat(weight_rast, se_disc(3))")
```

`tophat()`'s definition (`k - opening(k)`) is exactly that: a
real-valued residual (here, exactly `20` at the bump, matching its own
height above the plateau) highlighting where the raster's own values
exceed its own opening, not a binarized approximation of one. On a 0/1
binary raster it reduces exactly to the familiar `k AND NOT opening(k)` -
not approximately, since `opening(x) <= x` pointwise whenever the
structuring element includes its own centre (true for every named
shortcut at every radius). `bottomhat()` is the mirror image - a small
*dip* below the plateau, filled back in by `closing()`:

```{r continuous-bottomhat, fig.width = 10, fig.height = 4}
dip <- ifel(sqrt((gcx + 10)^2 + (gcy - 5)^2) <= 2, -20, 0)
weight_rast2 <- plateau + dip

weight_bh <- bottomhat(weight_rast2, se_disc(3))
par(mfrow = c(1, 2))
plot(weight_rast2, col = hcl.colors(50, "viridis"), axes = FALSE, main = "weight_rast2 (plateau + one dip)")
plot(weight_bh, col = hcl.colors(50, "viridis"), axes = FALSE, main = "bottomhat(weight_rast2, se_disc(3))")
```

# Categorical rasters

Categorical rasters (`terra::is.factor()` - discrete, unordered class
codes, e.g. a land-cover classification) are a genuinely different case. Taking the local
min/max of the raw CODE NUMBERS would only mean something spatially if
those codes happened to carry a real order, and for nominal categories
they don't - there's no meaningful sense in which land-cover code `1`
(say, forest) is "less than" code `3` (say, urban). So `erode()`/`dilate()`
switch to standard label-image morphology instead:

- **`erode()`**: a cell keeps its own label only if every cell in its
  neighbourhood shares that exact label; otherwise it becomes unlabelled
  (`NA`). Equivalent to eroding each class's own binary mask separately
  and taking the union - classes are mutually exclusive, so there is  never
  an overlap to resolve.
- **`dilate()`**: an already-labelled cell is left untouched; an
  unlabelled (`NA`) cell is filled with the majority label among its
  labelled neighbours. Restricting the fill to `NA` cells only is what
  keeps this monotonic (the defining property of dilation - it only ever
  grows into background, never reassigns existing foreground). A blind
  majority filter applied everywhere would instead be a smoothing
  operation, relabelling already-correct interior cells near a strong
  neighbouring majority.

One landcover raster with a stray class-2 cell poking into class-1
territory AND a one-cell interior gap - far enough from the class
boundary that it's a genuine gap, not just an eroded edge - shows all
four operators at once. The circles mark the two cells of interest, in
the same spatial position in every panel:

```{r categorical-setup, fig.width = 12, fig.height = 4}
n <- 25
landcover <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
values(landcover) <- 1
landcover[, 19:25] <- 2       # a solid class-2 block on the right
landcover[12, 18] <- 2        # a stray class-2 cell poking into class 1
landcover[12, 6] <- NA        # an interior gap, well away from the class boundary
landcover <- as.factor(landcover)
k <- se_box(1)

spike_xy <- xyFromCell(landcover, cellFromRowCol(landcover, 12, 18))
notch_xy <- xyFromCell(landcover, cellFromRowCol(landcover, 12, 6))

par(mfrow = c(1, 3))
plot(landcover, col = c("forestgreen", "steelblue"), axes = FALSE, main = "landcover (spike + gap)")
points(rbind(spike_xy, notch_xy), cex = 4, col = "red", lwd = 3)

plot(tophat(landcover, k), col = c("white", "firebrick"), axes = FALSE, main = "tophat(landcover, se_box(1))")
points(spike_xy, cex = 4, col = "black", lwd = 2)

plot(bottomhat(landcover, k), col = c("white", "firebrick"), axes = FALSE, main = "bottomhat(landcover, se_box(1))")
points(notch_xy, cex = 4, col = "black", lwd = 2)
```

`opening()` (erode then dilate) removes the stray cell entirely - too
thin to survive erosion, then not re-grown by dilation since it isn't
adjacent to the rest of class 2 - and `tophat()` (`!is.na(mask) &
is.na(opening(mask))`, the label-image analogue of `mask AND NOT
opening(mask)`) flags exactly that one cell, nothing else. The gap works
the same way in the other direction: `closing()` (dilate then erode)
fills it back in from its class-1 neighbours, and `bottomhat()`
(`is.na(mask) & !is.na(closing(mask))`) flags exactly that cell.
Subtracting category codes the way the continuous residual does would be
meaningless for either direction (what would "class 1 minus class 2"
even mean?) - `tophat()`/`bottomhat()` return a boolean flag for
categorical input, `0`/`1` but not itself categorical, rather than a
numeric residual.

To erode or dilate a single class on its own terms - rather than the
symmetric, every-class-at-once behaviour above - convert that class to
its own binary mask first (e.g. `landcover == 2`) and run the operator on
that instead.

# What each operator means, by raster type

|              | binary (0/1)                          | continuous                              | categorical                                  |
|--------------|----------------------------------------|------------------------------------------|-----------------------------------------------|
| `erode()`    | shrink the shape by one SE radius      | local minimum filter (grayscale erosion) | keep a label only if the whole neighbourhood agrees; else `NA` |
| `dilate()`   | grow the shape by one SE radius        | local maximum filter (grayscale dilation)| fill `NA` cells by majority vote; labelled cells untouched |
| `opening()`  | remove small bright features            | grayscale opening (erode then dilate)     | remove small/thin single-class regions        |
| `closing()`  | fill small dark gaps                    | grayscale closing (dilate then erode)     | fill small gaps within a class region         |
| `tophat()`   | `mask AND NOT opening(mask)`, `0`/`1`  | `mask - opening(mask)`, real-valued       | `1` where a label was removed by `opening()`  |
| `bottomhat()`| `closing(mask) AND NOT mask`, `0`/`1`  | `closing(mask) - mask`, real-valued       | `1` where a gap was filled by `closing()`     |

The first two columns are a genuine generalization: grayscale morphology
(local min/max) restricted to exactly two values reproduces binary
morphology exactly, not approximately - continuous *contains* binary as
a special case. However Categorical rasters are different. Operators have to be has to be a different rule built on a different assumption (every class treated symmetrically, none privileged as
background). This might be particularly problematic if a binary raster is converted to a categorical. 
Looks can be deceiving, if we only pay attention to category 1.

```{r}
disk_factor <- as.factor(noisy)
r1 <- opening(disk_factor, se_diamond(2))
r2 <- opening(noisy, se_diamond(2))

par(mfrow = c(1, 3))
plot(disk_factor, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "Categorical raster", colNA = 'red')
plot(r1, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "Opening (Cat)", colNA = "red")
plot(r2, col = c("white", "steelblue"), legend = FALSE, axes = FALSE, main = "Opening (Bin)", colNA = "red")

```


# Practical notes

**`NA` = hole**, consistently: an `NA` cell in a binary or continuous
mask is treated as background (the same as `0`) for `erode()`/`dilate()`;
in a categorical raster, `NA` is the *only* thing treated as background -
every other class code, including `0` if it's one of the raster's own
levels, is a real, persisting label (see `?gridmorph` for how this
compares to the index functions' own, deliberately different, "`0` is
always a hole" convention).

**The raster's own true edge is handled correctly**, not silently
treated as either "definitely background" or "definitely not a
boundary": a shape reaching the edge of the raster erodes there exactly
as it would next to an interior `NA` hole, and a categorical raster's own
edge behaves the same way an adjacent unknown region would (verified
directly against `terra::focal()`'s own edge-padding defaults, which
would otherwise under- or over-erode a shape that touches the raster's
extent).

**Built on `terra::focal()`, chunk-aware on disk-backed rasters the same
way `terra::distance()`/`terra::patches()` already are** - not `mmand`,
despite `mmand::erode()`/`dilate()` being somewhat faster per call at
sizes that already fit in memory. `mmand` operates on an in-memory array,
full stop; there's no way to run it on a raster too large to fit in
memory, which matters more for this package than the speed difference.

# See also

`vignette("a-basic-usage")` covers the thirteen shape indices and how
`weighted` works for them; `vignette("c-comparison-with-shapeindices")`
compares `gridmorph` against its vector-based sibling package on
accuracy, speed, and memory.
