---
title: "4. Resolution-Matched References"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
vignette: >
  %\VignetteIndexEntry{4. Resolution-Matched References}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center")
```

```{r load}
library(gridmorph)
library(terra)
```

Four of `gridmorph`'s thirteen indices - `gm_depth_index()`,
`gm_polsby_popper_index()`, `gm_reock_index()`, `gm_detour_index()` - work
by comparing something measured on the raster against a reference value
for a perfect circle. In each of these four, that reference used to be a
plain closed-form formula: the textbook constant a smooth, infinitely
precise circle would give. At any finite raster resolution, that turned
out to be the wrong number to compare against - even a perfectly
rasterized disk scored measurably below 1. This vignette shows the
symptom, explains why it happens, and describes the fix now built into
all four functions.

# The symptom

A disk is the one shape every one of these four indices should score at
(or extremely close to) 1 - that is each index's own definition of
"optimal". Reconstructing what each function's raw, pre-reference-fix
formula would have given - using only the geometric quantities each
function already returns (`area`, `perimeter`, `mbc_area`,
`hull_perimeter`) - shows all four undershooting 1 well outside any
reasonable rounding:

```{r symptom}
make_disk <- function(n = 81, radius = 30) {
  r <- rast(nrows = n, ncols = n, xmin = 0, xmax = n, ymin = 0, ymax = n, crs = "local")
  cx <- init(r, "x") - n / 2; cy <- init(r, "y") - n / 2
  ifel(sqrt(cx^2 + cy^2) <= radius, 1, 0)
}
disk <- make_disk()

depth   <- gm_depth_index(disk)
pp      <- gm_polsby_popper_index(disk)
reock   <- gm_reock_index(disk)
detour  <- gm_detour_index(disk)

raw_scores <- c(
  depth          = depth$mean_depth / (sqrt(depth$area / pi) / 3),
  polsby_popper  = 4 * pi * pp$area / pp$perimeter^2,
  reock          = reock$area / reock$mbc_area,
  detour         = 2 * sqrt(pi * detour$area) / detour$hull_perimeter
)
knitr::kable(format = "html", round(raw_scores, 3), col.names = "raw score (a perfect disk should give 1)")
```

Comparing this to what these four functions actually return today shows
the fix in effect - every one lands close to 1:

```{r fixed}
fixed_scores <- c(depth = depth$index, polsby_popper = pp$index,
                   reock = reock$index, detour = detour$index)
knitr::kable(format = "html", round(fixed_scores, 3), col.names = "gridmorph's actual index")
```

# Why a perfect disk doesn't score exactly 1

The four functions fall into two different mechanisms, both boiling down
to the same root cause: **a raster is a set of finite cells, not a
continuous shape**, so anything measured from it - a distance, a
perimeter, an area by pixel count - is only ever an approximation of the
matching continuous quantity, and that approximation doesn't hit the
continuous formula's own constant exactly. Zooming in on a small
rasterized disk against the true circle it approximates makes this
concrete - the shape these four indices actually measure is the blue
stair-step region, not the smooth red curve:

```{r fig-raster-vs-circle, echo = FALSE, fig.width = 5.5, fig.height = 5.5}
n <- 16; radius <- 6
cx <- n / 2; cy <- n / 2
xy <- expand.grid(i = 1:n, j = 1:n)
xy$x <- xy$i - 0.5; xy$y <- xy$j - 0.5
xy$valid <- sqrt((xy$x - cx)^2 + (xy$y - cy)^2) <= radius

plot(NA, xlim = c(0, n), ylim = c(0, n), asp = 1, axes = FALSE, xlab = "", ylab = "",
     main = "A rasterized disk vs. the true circle it approximates")
for (i in seq_len(nrow(xy))) {
  rect(xy$i[i] - 1, xy$j[i] - 1, xy$i[i], xy$j[i],
       col = if (xy$valid[i]) "#8fb8de" else "white", border = "grey85")
}
theta <- seq(0, 2 * pi, length.out = 300)
lines(cx + radius * cos(theta), cy + radius * sin(theta), col = "firebrick", lwd = 2.5)
points(xy$x[xy$valid], xy$y[xy$valid], pch = 16, cex = 0.5, col = "grey25")
legend("topright", legend = c("true continuous circle", "valid cell centre"),
       col = c("firebrick", "grey25"), lty = c(1, NA), pch = c(NA, 16), lwd = c(2.5, NA),
       bty = "n", cex = 0.9)
```

**`gm_depth_index()`**: `mean_depth` comes from `terra::distance()`,
which measures from a cell's own *centre* to the nearest invalid cell's
*centre* - not to the true boundary curve. That's a systematically larger
quantity than true distance-to-boundary, most so right at the boundary
and decaying (but never fully vanishing at any finite resolution) toward
the interior. Zooming in on one boundary cell shows the gap directly - the
true distance to the circle (green) is shorter than the distance
`terra::distance()` actually measures, to the centre of the nearest
invalid cell (orange, dashed):

```{r fig-depth-mechanism, echo = FALSE, fig.width = 7.5, fig.height = 5.2}
row_y <- cy + 0.5
row <- xy[xy$valid & abs(xy$y - row_y) < 1e-6, ]
pick <- row[which.max(row$x), ]
true_pt <- c(cx + radius, pick$y)
inv <- xy[!xy$valid, ]
inv$dist_to_pick <- sqrt((inv$x - pick$x)^2 + (inv$y - pick$y)^2)
nearest_inv <- inv[which.min(inv$dist_to_pick), ]

xlim <- c(pick$x - 3.5, pick$x + 3); ylim <- c(pick$y - 3, pick$y + 3.5)
plot(NA, xlim = xlim, ylim = ylim, asp = 1, axes = FALSE, xlab = "", ylab = "",
     main = "gm_depth_index(): what terra::distance() measures vs. true depth")
for (i in seq_len(nrow(xy))) {
  if (xy$i[i] - 1 < xlim[1] - 1 || xy$i[i] > xlim[2] + 1) next
  if (xy$j[i] - 1 < ylim[1] - 1 || xy$j[i] > ylim[2] + 1) next
  rect(xy$i[i] - 1, xy$j[i] - 1, xy$i[i], xy$j[i],
       col = if (xy$valid[i]) "#8fb8de" else "white", border = "grey85")
}
rect(pick$i - 1, pick$j - 1, pick$i, pick$j, col = "#f4c66b", border = "grey60")
theta_r <- seq(-pi / 2, pi / 2, length.out = 200)
lines(cx + radius * cos(theta_r), cy + radius * sin(theta_r), col = "firebrick", lwd = 2.5)
off <- 0.16
arrows(pick$x, pick$y + off, true_pt[1], pick$y + off, col = "#1a7a1a", lwd = 3, length = 0.13)
arrows(pick$x, pick$y - off, nearest_inv$x, pick$y - off, col = "#b3590a", lwd = 3, length = 0.13, lty = 2)
points(pick$x, pick$y, pch = 16, cex = 1.3)
legend("bottomleft", legend = c("true distance to the circle", "terra::distance() measures this"),
       col = c("#1a7a1a", "#b3590a"), lty = c(1, 2), lwd = 3, bty = "n", cex = 0.85)
```

**`gm_polsby_popper_index()`**: perimeter comes from marching squares
(`terra::as.contour()`), a technique built to reconstruct a smooth
boundary from a *smoothly-sampled* field. A binary raster mask was never
smoothly sampled in the first place - there's no genuine sub-cell signal
for it to interpolate - so the reconstructed perimeter carries a real,
persistent overestimate that does not shrink away at finer resolutions,
on top of substantial noise depending on exactly how the shape's boundary
happens to sit against the grid. The left panel shows the reconstructed
contour (dark blue) cutting corners against the true circle in some
places and bulging past it in others; the right panel shows twelve disks
of the *same* radius, each centred a random fraction of a cell
differently, and how much that alone moves the (already
resolution-matched) index:

```{r fig-pp-mechanism, echo = FALSE, fig.width = 10, fig.height = 5}
n2 <- 20; radius2 <- 7
r2 <- rast(nrows = n2, ncols = n2, xmin = 0, xmax = n2, ymin = 0, ymax = n2, crs = "local")
cxr <- init(r2, "x") - n2 / 2; cyr <- init(r2, "y") - n2 / 2
mask2 <- ifel(sqrt(cxr^2 + cyr^2) <= radius2, 1, 0)
padded2 <- extend(mask2, 1, fill = 0)
cont2 <- as.contour(padded2, levels = 0.5)
cont_xy <- geom(cont2)

par(mfrow = c(1, 2))
cx2 <- n2 / 2; cy2 <- n2 / 2
xy2 <- expand.grid(i = 1:n2, j = 1:n2)
xy2$x <- xy2$i - 0.5; xy2$y <- xy2$j - 0.5
xy2$valid <- sqrt((xy2$x - cx2)^2 + (xy2$y - cy2)^2) <= radius2

plot(NA, xlim = c(3, n2 - 3), ylim = c(3, n2 - 3), asp = 1, axes = FALSE, xlab = "", ylab = "",
     main = "marching-squares perimeter vs. the true circle")
for (i in seq_len(nrow(xy2))) {
  rect(xy2$i[i] - 1, xy2$j[i] - 1, xy2$i[i], xy2$j[i],
       col = if (xy2$valid[i]) "#8fb8de" else "white", border = "grey85")
}
theta2 <- seq(0, 2 * pi, length.out = 300)
lines(cx2 + radius2 * cos(theta2), cy2 + radius2 * sin(theta2), col = "firebrick", lwd = 2)
lines(cont_xy[, "x"], cont_xy[, "y"], col = "#1a3d6b", lwd = 2.5)
legend("topright", legend = c("true circle", "marching-squares contour"),
       col = c("firebrick", "#1a3d6b"), lwd = c(2, 2.5), bty = "n", cex = 0.85)

set.seed(11)
n_ph <- 12
offs <- data.frame(dx = runif(n_ph, -0.45, 0.45), dy = runif(n_ph, -0.45, 0.45))
pp_scores <- numeric(n_ph)
for (k in seq_len(n_ph)) {
  rr <- rast(nrows = n2, ncols = n2, xmin = 0, xmax = n2, ymin = 0, ymax = n2, crs = "local")
  xyk <- xyFromCell(rr, 1:ncell(rr))
  dk <- sqrt((xyk[, 1] - (n2 / 2 + offs$dx[k]))^2 + (xyk[, 2] - (n2 / 2 + offs$dy[k]))^2)
  values(rr) <- ifelse(dk <= radius2, 1, NA)
  pp_scores[k] <- gm_polsby_popper_index(rr)$index
}
plot(seq_len(n_ph), pp_scores, pch = 16, cex = 1.4, col = "#1a3d6b",
     xlab = "12 same-radius disks, different sub-cell phase", ylab = "gm_polsby_popper_index()$index",
     main = "residual phase noise, even after the fix", ylim = c(0.9, 1.05), xaxt = "n")
abline(h = 1, col = "firebrick", lty = 2, lwd = 2)
axis(1, at = seq_len(n_ph), labels = FALSE)
```

**`gm_reock_index()` / `gm_detour_index()`**: both compare a pixel-COUNT
area (or a raster-derived convex hull) against an EXACTLY-computed
reference (the minimum bounding circle's area, or a perfect circle's
analytical perimeter). Pixel counting alone always slightly undershoots a
shape's true continuous area near its own boundary, while the exact
reference carries no matching imprecision - so the two sides of the ratio
are measured by two different conventions that don't cancel by
themselves. Under centre-based rasterization, a boundary cell's own far
corner can sit outside the true circle - the hull built from those
corners (dark blue) visibly bulges past it:

```{r fig-hull-mechanism, echo = FALSE, fig.width = 7, fig.height = 5.8}
corners <- do.call(rbind, lapply(seq_len(nrow(xy)), function(k) {
  if (!xy$valid[k]) return(NULL)
  cxs <- c(xy$i[k] - 1, xy$i[k]); cys <- c(xy$j[k] - 1, xy$j[k])
  cc <- expand.grid(x = cxs, y = cys)
  cc$d <- sqrt((cc$x - cx)^2 + (cc$y - cy)^2)
  cc[which.max(cc$d), ]
}))
hull_idx <- chull(corners$x, corners$y)
hull_pts <- corners[c(hull_idx, hull_idx[1]), ]

plot(NA, xlim = c(0, n), ylim = c(0, n + 3), asp = 1, axes = FALSE, xlab = "", ylab = "",
     main = "a boundary cell's own outer corner sits outside the true circle")
for (i in seq_len(nrow(xy))) {
  rect(xy$i[i] - 1, xy$j[i] - 1, xy$i[i], xy$j[i],
       col = if (xy$valid[i]) "#8fb8de" else "white", border = "grey85")
}
lines(cx + radius * cos(theta), cy + radius * sin(theta), col = "firebrick", lwd = 2.5)
lines(hull_pts$x, hull_pts$y, col = "#1a3d6b", lwd = 2.5)
points(corners$x, corners$y, pch = 16, cex = 0.5, col = "grey40")
pick2 <- xy[xy$valid & abs(xy$y - (cy + 0.5)) < 1e-6, ]
pick2 <- pick2[which.max(pick2$x), ]
rect(pick2$i - 1, pick2$j - 1, pick2$i, pick2$j, col = "#f4c66b", border = "grey50", lwd = 1.5)
points(pick2$i, pick2$j, pch = 16, cex = 1.2, col = "#b3590a")
legend("topleft", legend = c("true circle", "hull of cell corners", "a boundary cell's own outer corner"),
       col = c("firebrick", "#1a3d6b", "#b3590a"), lwd = c(2.5, 2.5, NA), pch = c(NA, NA, 16),
       bty = "n", cex = 0.85)
```

# The fix: measure the reference the same way

The common fix, used identically by all four: instead of comparing the
shape's raster-measured score against a pure formula, build an *actual*
rasterized disk at the input's own cell size and area, run it through the
exact same measurement pipeline the shape itself just went through, and
divide by that instead. Since both sides of the ratio now carry the same
measurement artifacts, those artifacts cancel out rather than leaking
into the final index:

```{r ref-index}
c(polsby_popper_ref = pp$ref_index, reock_ref = reock$ref_index, detour_ref = detour$ref_index)
```

Each of these `ref_index` values is exactly what a same-resolution
rasterized disk itself scores on the *raw*, uncorrected formula - the
same numbers as the "raw score" table above, just for a disk built to
match the input's own resolution rather than the input itself.
`gm_depth_index()`'s own reference is the analogous quantity - a
same-resolution disk's own mean distance-to-boundary - returned as
`ref_depth` rather than `ref_index`, since it's a physical distance, not
a dimensionless score.

One index needs a further refinement. `gm_polsby_popper_index()`'s
marching-squares perimeter is sensitive enough to exactly where a shape's
boundary falls against the pixel grid that a *single* reference disk
isn't a reliable stand-in - two disks of the *same radius*, centred a
fraction of a cell apart, can give visibly different raw scores. The left
panel below overlays two such disks directly: one centred on a cell
centre (solid dark outline), one shifted `0.45` cells diagonally (dashed
orange outline) - the blue-shaded cells are only in the first disk, the
orange-shaded cells only in the second. That's the entire source of the
difference between their raw scores. The right panel shows where those
two scores (triangles) sit among the full 25-point grid of sub-cell
offsets `gm_polsby_popper_index()` actually averages over: scattered on
both sides of the mean (the dark blue line) - no single one of those 25
points would make a reliable reference on its own, but their mean is
stable enough to use:

```{r fig-phase-averaging, echo = FALSE, fig.width = 11, fig.height = 4.8}
make_disk_at <- function(nn, rr, dx, dy) {
  rst <- rast(nrows = nn, ncols = nn, xmin = 0, xmax = nn, ymin = 0, ymax = nn, crs = "local")
  xyk <- xyFromCell(rst, 1:ncell(rst))
  dd <- sqrt((xyk[, 1] - (nn / 2 + dx))^2 + (xyk[, 2] - (nn / 2 + dy))^2)
  values(rst) <- ifelse(dd <= rr, 1, NA)
  rst
}
raw_pp <- function(rst) {
  res <- gm_polsby_popper_index(rst)
  4 * pi * res$area / res$perimeter^2
}
contour_of <- function(rst) {
  valid <- !is.na(rst)
  padded <- extend(ifel(valid, 1, 0), 1, fill = 0)
  geom(as.contour(padded, levels = 0.5))
}

n3 <- 14; radius3 <- 5  # coarser than the figures above, so a 0.45-cell phase shift is clearly visible
offsets25 <- seq(-0.45, 0.45, length.out = 5)
grid_off <- expand.grid(dx = offsets25, dy = offsets25)
raw_grid <- vapply(seq_len(nrow(grid_off)), function(k) raw_pp(make_disk_at(n3, radius3, grid_off$dx[k], grid_off$dy[k])), numeric(1))
avg_ref <- mean(raw_grid)
idx1 <- which(grid_off$dx == 0 & grid_off$dy == 0)
idx2 <- which.max(grid_off$dx + grid_off$dy)
disk_a <- make_disk_at(n3, radius3, grid_off$dx[idx1], grid_off$dy[idx1])
disk_b <- make_disk_at(n3, radius3, grid_off$dx[idx2], grid_off$dy[idx2])
cont_a <- contour_of(disk_a); cont_b <- contour_of(disk_b)

par(mfrow = c(1, 2))
xy_a <- xyFromCell(disk_a, 1:ncell(disk_a)); v_a <- as.vector(values(!is.na(disk_a)))
xy_b <- xyFromCell(disk_b, 1:ncell(disk_b)); v_b <- as.vector(values(!is.na(disk_b)))
only_a <- v_a & !v_b; only_b <- v_b & !v_a; both_ab <- v_a & v_b

plot(NA, xlim = c(1, n3 - 1), ylim = c(1, n3 - 1), asp = 1, axes = FALSE, xlab = "", ylab = "",
     main = "same radius, different sub-cell phase")
for (i in which(both_ab)) rect(xy_a[i, 1] - 0.5, xy_a[i, 2] - 0.5, xy_a[i, 1] + 0.5, xy_a[i, 2] + 0.5, col = "#c9d9ea", border = "grey85")
for (i in which(only_a)) rect(xy_a[i, 1] - 0.5, xy_a[i, 2] - 0.5, xy_a[i, 1] + 0.5, xy_a[i, 2] + 0.5, col = "#6fa8dc", border = "grey60")
for (i in which(only_b)) rect(xy_b[i, 1] - 0.5, xy_b[i, 2] - 0.5, xy_b[i, 1] + 0.5, xy_b[i, 2] + 0.5, col = "#f4b183", border = "grey60")
lines(cont_a[, "x"], cont_a[, "y"], col = "#1a3d6b", lwd = 3)
lines(cont_b[, "x"], cont_b[, "y"], col = "#b3590a", lwd = 3, lty = 2)
legend("topleft", legend = c(sprintf("centred on a cell centre (raw = %.3f)", raw_grid[idx1]),
                              sprintf("shifted 0.45 cells diagonally (raw = %.3f)", raw_grid[idx2]),
                              "cells only the first disk includes", "cells only the second disk includes"),
       col = c("#1a3d6b", "#b3590a", NA, NA), lty = c(1, 2, NA, NA), lwd = c(3, 3, NA, NA),
       fill = c(NA, NA, "#6fa8dc", "#f4b183"), border = NA, bty = "n", cex = 0.8)

plot(seq_along(raw_grid), sort(raw_grid), pch = 16, col = "grey50", cex = 1.4,
     xlab = "the 25 sub-cell offsets, sorted by score", ylab = "raw polsby-popper score",
     main = "single offsets are noisy; the mean isn't", xaxt = "n")
axis(1, at = seq_along(raw_grid), labels = FALSE)
abline(h = avg_ref, col = "#1a3d6b", lwd = 3)
r1_rank <- match(raw_grid[idx1], sort(raw_grid))
r2_rank <- match(raw_grid[idx2], sort(raw_grid))
points(r1_rank, raw_grid[idx1], pch = 17, cex = 2, col = "#1a3d6b")
points(r2_rank, raw_grid[idx2], pch = 17, cex = 2, col = "#b3590a")
text(r1_rank, raw_grid[idx1], "centred", pos = 3, cex = 0.9, col = "#1a3d6b")
text(r2_rank, raw_grid[idx2], "shifted 0.45", pos = 4, cex = 0.9, col = "#b3590a")
legend("bottomright", legend = sprintf("mean across all 25 = ref_index = %.3f", avg_ref),
       col = "#1a3d6b", lwd = 3, bty = "n", cex = 0.9)
```

That reference is instead averaged over a grid of 25 sub-cell offsets,
which was enough to keep the corrected index within about half a percent
of 1 across every resolution checked. `gm_reock_index()` and
`gm_detour_index()` don't need this: their own raw scores were checked
directly and shown to vary by only a few thousandths across the same
offsets, so a single reference disk is already a reliable comparison.

None of these four functions expose a way to opt back into the old,
purely-analytical comparison - the number it used to give was not a
different valid choice, it was measurably the wrong one for a rasterized
shape at any finite resolution.

Running all four functions on disks of increasing radius shows what the
fix buys, and where it doesn't: `gm_depth_index()`/`gm_reock_index()`/
`gm_detour_index()`'s raw scores (orange) converge toward 1 only slowly,
needing a genuinely large radius to get close - the resolution-matched
`index` (blue) is already tight at every radius shown, including the
smallest. `gm_polsby_popper_index()`'s raw score doesn't converge toward 1
at all, even at radius 90 - the non-vanishing bias described above - while
its own resolution-matched `index` stays close to 1 throughout, with the
visible wobble being exactly the residual phase noise from the figure
above:

```{r fig-convergence, echo = FALSE, fig.width = 10, fig.height = 6}
radii <- c(10, 15, 20, 30, 45, 60, 90)
conv <- data.frame()
for (rad in radii) {
  nn <- ceiling(rad * 2.7)
  dk <- make_disk(nn, rad)
  d_res <- gm_depth_index(dk); pp_res <- gm_polsby_popper_index(dk)
  reock_res <- gm_reock_index(dk); detour_res <- gm_detour_index(dk)
  raw <- c(depth = d_res$mean_depth / (sqrt(d_res$area / pi) / 3),
           polsby_popper = 4 * pi * pp_res$area / pp_res$perimeter^2,
           reock = reock_res$area / reock_res$mbc_area,
           detour = 2 * sqrt(pi * detour_res$area) / detour_res$hull_perimeter)
  fixed <- c(depth = d_res$index, polsby_popper = pp_res$index,
             reock = reock_res$index, detour = detour_res$index)
  conv <- rbind(conv, data.frame(radius = rad, index = names(raw), raw = raw, fixed = fixed))
}

par(mfrow = c(2, 2), mar = c(4.2, 4.2, 3, 1))
idx_names <- c("depth", "polsby_popper", "reock", "detour")
titles <- c("gm_depth_index()", "gm_polsby_popper_index()", "gm_reock_index()", "gm_detour_index()")
for (k in seq_along(idx_names)) {
  sub <- conv[conv$index == idx_names[k], ]
  yr <- range(c(sub$raw, sub$fixed, 1))
  plot(sub$radius, sub$raw, type = "b", pch = 17, col = "#b3590a", lwd = 2, cex = 1.2,
       ylim = yr, xlab = "disk radius (cells)", ylab = "index value", main = titles[k], log = "x")
  lines(sub$radius, sub$fixed, type = "b", pch = 16, col = "#1a3d6b", lwd = 2, cex = 1.2)
  abline(h = 1, col = "grey50", lty = 3)
  if (k == 1) legend("bottomleft", legend = c("raw (vs. analytical formula)", "gridmorph's actual index"),
                      col = c("#b3590a", "#1a3d6b"), pch = c(17, 16), lwd = 2, bty = "n", cex = 0.8)
}
```

# What this doesn't cover

`gm_hull_ratio_index()` (`area / hull_area`) and `gm_exchange_index()`
(share of a shape's own area inside a reference circle at its centroid)
are not part of this fix, and a rasterized disk still scores a little
below 1 on both:

```{r not-covered}
c(hull_ratio = gm_hull_ratio_index(disk)$index, exchange = gm_exchange_index(disk)$index)
```

Both are self-referential - nowhere in either formula does an external
circle formula appear to be measured against. `gm_hull_ratio_index()`
compares a shape to its *own* convex hull; `gm_exchange_index()` compares
a shape to a circle built from the shape's *own* area and centroid. A
rasterized disk's stair-step boundary genuinely isn't perfectly convex at
the pixel level - real, tiny notches exist along it, the same way a
coastline traced at high enough zoom genuinely isn't a smooth curve. A
same-resolution reference disk would show the same small gap, not erase
it, because there is nothing measured incorrectly here to begin with.
Both converge toward 1 as resolution increases, the ordinary way any
raster-native measurement converges toward its continuous limit.
